{\fontfamily{ptm}\selectfont Nesterov Type Momentum Methods}

### 2024-06-06 Notebook LaTEX/Momentum Accelerations/main.tex
>w< fixes and updates

\section{The proximal point method with convexity}
    This section reviews the convex case's Proximal point method (PPM) analysis and generalizes the theories to approximated PPM. 
    \subsection{Literature reviews}
        Rockafellar \cite{rockafellar_monotone_1976} pioneered the analysis of the proximal point method in the convex case. 
        He developed the analysis in the context of maximal monotone operators in Hilbert spaces. 
        Applications in convex optimizations are covered. 
        Using his theorems appropriately requires some opportunities, realizations, and characterizations of assumptions (A), (B) in his paper in the context of the applications.

@@

        \begin{theorem}
            With $f$ being $\beta\ge 0$ strongly convex and $x_{t + 1} = \hprox_{\eta_{t + 1}f}$ generated by PPM. 

@@

        \begin{theorem}[Convergence Rate of PPM]\label{thm:ppm_convergence_rate}
            The convergence rate of PPM applied to $f$, closed, convex proper, we have the convergence rate of the function value: 
            $$
            f(x_T) - f(x_*) \le O\left(\left(\sum_{i=1}^{T}\eta_t\right)^{-1}\right). 
            $$
            Where $x_*$ is the minimizer of $f$. 
        \end{theorem}
        \begin{proof}
            With $\Delta_t = f(x_t) - f(x_*), \Upsilon_t = \sum_{i = 1}^{t}\eta_i$ so $\Phi_t = \Upsilon_t\Delta_t + \frac{1}{2}\Vert x_t - x_*\Vert^2$ by consideration $u = x_*$, invoking previous theorem and do
            \begin{align*}
                \Upsilon_T\Delta_T \le \Phi_T 
                &\le 
                \Phi_0 = \frac{1}{2}\Vert x_0 - x_*\Vert^2 
                \\
                \implies \Delta_T 
                &\le 
                \frac{1}{2\Upsilon_T} \Vert x_0 - x_*\Vert^2. 
            \end{align*}
        \end{proof}
        \begin{remark}
            With the same choice of the sequence $(\eta_t)_{t \in \NN}$, convergence of the PPM method of a strongly convex function is faster.     
        \end{remark}

@@

        \begin{example}[The fundamental proximal gradient lemma]
            The fundamental proximal gradient lemma was used heavily in the literature to derive convergence results in the convex case. 
            The "fundamental proximal gradient lemma" originates from Beck's writings\cite[theorem 10.16]{beck_first-order_nodate}. 
            We demonstrate in this example that it's a consequence of \hyperref[thm:ppm_breg_descent_ineq]{theorem \ref*{thm:ppm_breg_descent_ineq}}. 
            \par
            With $f = g + h$, $h$ convex, lsc, $g$ be $L$-Lipschitz smooth, then for all $y\in \RR^n$, $x\in \RR^n$, $y^+ := \hprox_{L^{-1}h}(y - L^{-1}\nabla (y))$ satisfies: 
            \begin{align*}
                f(x) - f(y^+) \ge \frac{L}{2}\Vert x - y^+\Vert^2 - \frac{L}{2}\Vert x - y\Vert^2 + D_g(x, y).
            \end{align*}
            A similar analysis as \hyperref[thm:lower_approx_ppm_convergence]{ theorem \ref*{thm:lower_approx_ppm_convergence}} with \hyperref[thm:ppm_breg_descent_ineq]{theorem \ref*{thm:ppm_breg_descent_ineq}} obtains the same inquality. 
            With $\phi(y) = \eta(h(y) + g(x) + \langle \nabla g(x), y - x\rangle)$ as an lower bounding function of $f$. 
            Let $x^+ = \hprox_\phi(x)$, then for all $u$:
            \begin{align*}
                & \phi(u) + \frac{1}{2}\Vert u - x\Vert^2 - \phi(x^+) - \frac{1}{2}\Vert x^+ - x\Vert^2 
                \ge \frac{1}{2}\Vert x^+ - u\Vert^2, 
                \\
                \implies &
                \eta\underbrace{(h(u) + g(x) + \langle \nabla g(x), u - x\rangle + \frac{L}{2}\Vert u - x\Vert^2)}_{\ge \phi(u)} 
                - \eta \underbrace{f(x^+)}_{\le \phi(x^+)} - \frac{1}{2}\Vert x - x^+\Vert^2 
                \\
                &\quad  
                + \left(
                    \frac{1}{2} - \frac{\eta L}{2}
                \right)\Vert u - x\Vert^2 
                \ge 
                \frac{1}{2}\Vert x^+ - u\Vert^2 
                \\
                \iff & 
                f(u) + g(x) - g(u) + 
                \langle \nabla g(x), u -x\rangle 
                - f(x^+) + \frac{L}{2}\Vert u - x\Vert^2 - \frac{1}{2\eta}\Vert x - x^+\Vert^2
                \\
                & \quad 
                + 
                \left(
                    \frac{1}{2\eta} - \frac{L}{2}
                \right)\Vert u - x\Vert^2 
                \ge 
                \frac{1}{2\eta}\Vert x^+ - u\Vert^2 
                \\
                \iff 
                & f(u) - f(x^+) - D_g(u, x)
                + \frac{1}{2\eta} \Vert u - x\Vert^2 - \frac{1}{2\eta}\Vert x^+ - x\Vert^2
                \ge 
                \frac{1}{2\eta} \Vert x^+ - u\Vert^2. 
            \end{align*}
            Removing the negative term $-1/2\eta \Vert x - x^+\Vert^2$ makes LHS larger, establishing the fundamental proximal gradient lemma. 
        \end{example}

@@

    \subsection{Varieties of Nesterov accelerated gradient}
        In this section, we list different varieties of the Nesterov accelerated method. 
        We present these varieties generically because these algorithms' forms are of interest.  
        \begin{definition}[Accelearted Gradient Generic Original Form]
            Let $f$ be a $L$ Lipschitz smooth and $\mu\ge 0$ strongly convex function. 
            Choose $x_0$, $\gamma_0 > 0$, set $v_0 = x_0$, for iteration $k\ge 0$, it
            \begin{enumerate}
                \item[1.] computes $\alpha_k \in (0, 1)$ by solving $L\alpha_k^2 = (1 - \alpha_k)\gamma_k + \alpha_k \mu$; 
                \item[2.] sets $\gamma_{k + 1} = (1 - \alpha_k)\gamma_k + \alpha_k \mu$;
                \item[3.] chooses $y_k = (\gamma_k + \alpha_k \mu)(\alpha_k \gamma_k v_k + \gamma_{k + 1}x_k)$. Compute $f(y_k)$ and $\nabla f(y_k)$; 
                \item[4.] finds $x_{k + 1}$ such that $f(x_{k + 1}) \le f(y_k) - (2L)^{-1} \Vert \nabla f(y_k)\Vert^2$; 
                \item[5.] sets $v_{k + 1} = \gamma_{k+1}^{-1}((1 - \alpha_k)\gamma_kv_k + \alpha_k \mu y_k - \alpha_k \nabla f(y_k))$. 
            \end{enumerate}
        \end{definition}
        \begin{remark}
            This is in Nesterov's book \cite[(2.2.7)]{nesterov_lectures_2018}. 
            It is the most generic algorithm in his book about accelerated gradient method. 
            The genericity of the algorithm is provided by item 4., which is the a special case of the smooth descent lemma. 
        \end{remark}

        \begin{definition}[Accelerated Gradient Generic Triangular Form]

@@

        \begin{definition}[Accelerated Gradient Generic PPM Form]

>>>>

\section{The proximal point method with convexity}
    This section reviews the convex case's Proximal point method (PPM) analysis and generalizes the theories to approximated PPM. 
    \subsection{Convex PPM literature reviews}
        Rockafellar \cite{rockafellar_monotone_1976} pioneered the analysis of the proximal point method in the convex case. 
        He developed the analysis in the context of maximal monotone operators in Hilbert spaces. 
        Applications in convex optimizations are covered. 
        Using his theorems appropriately requires some opportunities, realizations, and characterizations of assumptions (A), (B) in his paper in the context of the applications. 
        \par\noindent
        In this section, we will use the result from Rockafellar that, if a monotone operator $A$ is $\beta$ strongly convex, then the resolvent operator $\mathcal J_A = [I + A]^{-1}$ is a $(1 + \beta)^{-1}$ Lipschitz operator, making $I - \mathcal J_A$ is a $1 - (1 + \beta)^{-1}$ a strongly monotone operator.

@@

        \begin{theorem}[PPM Lyapunov Function]\label{ppm_lyapunov}
            With $f$ being $\beta\ge 0$ convex (it's strongly convex if $\beta > 0$, else it's just convex) and $x_{t + 1} = \hprox_{\eta_{t + 1}f}$ generated by PPM. 

@@

        \begin{theorem}[Convergence Rate of PPM]\label{thm:ppm_convergence_rate}
            The convergence rate of PPM applied to $f$, closed, convex proper, we have the convergence rate of the function value: 
            $$
            f(x_T) - f(x_*) \le O\left(\left(\sum_{i=1}^{T}\eta_t\right)^{-1}\right). 
            $$
            Where $x_*$ is the minimizer of $f$. 
        \end{theorem}
        \begin{proof}
            With $\Delta_t = f(x_t) - f(x_*), \Upsilon_t = \sum_{i = 1}^{t}\eta_i$ so $\Phi_t = \Upsilon_t\Delta_t + \frac{1}{2}\Vert x_t - x_*\Vert^2$ by consideration $u = x_*$, invoking previous theorem and do
            \begin{align*}
                \Upsilon_T\Delta_T \le \Phi_T 
                &\le 
                \Phi_0 = \frac{1}{2}\Vert x_0 - x_*\Vert^2 
                \\
                \implies \Delta_T 
                &\le 
                \frac{1}{2\Upsilon_T} \Vert x_0 - x_*\Vert^2. 
            \end{align*}
        \end{proof}
        \begin{remark}
            With the same choice of the sequence $(\eta_t)_{t \in \NN}$, convergence of the PPM method of a strongly convex function is faster. 
            The above proof is the same for $\beta = 0$, or $\beta > 0$, because it didn't use the property that $\eta_{t + 1}f$ is a $\eta_{t + 1}\beta$ strongly convex function. 
        \end{remark}

        \begin{theorem}[PPM Strongly Convex Lyapunov Function]\label{ppm_lyapunov_scvx}
            With $f$ being $\beta > 0$ strogly convex, with $x_{t + 1} = \hprox_{\eta_{t + 1}f}(x_t)$, then $\Phi_t = f(x_t) - f(x_*)$ is a Lyapunov function satisfying: 
                f(x_{t + 1}) - f(x_*)
                -\frac{1}{2\eta_{t + 1}}(1 + \eta_{t + 1}\beta)\Vert x_{t + 1} - x_*\Vert^2
                & \quad 
                + \frac{1}{2\eta_{t + 1}}\left(
                    1 - (1 - (1 + \eta_{t + 1}\beta)^{-1})^2
                    \prod_{i = 1}^{t}
                        (1 + \eta_i \beta)^{-1}
                \right)\Vert x_0 - x_*\Vert^2, 
            hence a convergence rate of $\mathcal O\left(
                \prod_{i = 1}^{t}(1 + \eta_i \beta)^{-1}
            \right)$ is prominent. 
            With $\beta$-strongly convex $f$, let $\phi = \eta_{t + 1}f$, let $E_\phi(x)$ be: $\min_u \{\eta_{t + 1} f(u) + (1/2)\Vert u - x\Vert^2\}$ which is the Moreau envelope of $\phi$, let $P_\phi := [I + \eta_{t + 1}\partial f]^{-1}$. 
            For any $x_0$ let $x_{t + 1} = P_\phi(x_t)$, with $x_* \in \argmin_{x} \phi(x)$, we realize these consequences:
            \begin{enumerate}
                \item [1.] $I - P_\phi$ is a $1 - (1 + \eta_{t + 1}\beta)^{-1}$ strongly monotone operator;
                \item [2.] $\nabla E_\phi = I - P_\phi$, meaning that $E_\phi$ is $1 - (1 + \eta_{t + 1}\beta)^{-1}$ strongly convex;
                \item [3.] $P_\phi$ is a contractive mapping with constant $(1 + \beta\eta_{t + 1})^{-1}$. 
            \end{enumerate}
            Using strong convexity of the Moreau envelope it has 
                \frac{1}{2}\Vert \nabla E_\phi(x_t)\Vert^2 
                &\ge 
                (1 - (1 + \beta\eta_{t + 1})^{-1}) (E_\phi(x_t) - E_\phi(x_*))
                &\ge 
                (1 - (1 + \beta\eta_{t + 1})^{-1})\left(
                    \frac{1}{2}(1 - (1 + \beta\eta_{t + 1})^{-1})
                    \Vert x - x_*\Vert^2
                \frac{1}{2}\left(
                    1 - (1 + \eta_{t + 1}\beta)^{-1}
                \right)^2 \Vert x_t - x_*\Vert^2
            invoke PPM descent inequality on the function $\phi$, it has 
                E_\phi(x_{t + 1}) - \phi(x_*) 
                \frac{1}{2}\Vert x_t - x_*\Vert^2 
                -\frac{1 + \eta_{t + 1}\beta}{2}\Vert x_{t + 1} - x_*\Vert^2
                \phi(x_{t + 1}) - \phi(x_*)
                -\frac{1}{2}(1 + \eta_{t + 1}\beta)\Vert x_{t + 1} - x_*\Vert^2
                + \frac{1}{2}\Vert x_t - x_*\Vert^2
                \phi(x_{t + 1}) - \phi(x_*)
                -\frac{1}{2}(1 + \eta_{t + 1}\beta)\Vert x_{t + 1} - x_*\Vert^2
                - \frac{1}{2}(1 - (1 + \eta_{t + 1}\beta)^{-1})^2\Vert x_t - x_*\Vert^2
                + \frac{1}{2}\Vert x_t - x_*\Vert^2
                \phi(x_{t + 1}) - \phi(x_*)
                -\frac{1}{2}(1 + \eta_{t + 1}\beta)\Vert x_{t + 1} - x_*\Vert^2
                + \frac{1}{2}\left(
                    1 - (1 - (1 + \eta_{t + 1}\beta)^{-1})^2
                \Vert x_t - x_*\Vert^2
                \phi(x_{t + 1}) - \phi(x_*)
                -\frac{1}{2}(1 + \eta_{t + 1}\beta)\Vert x_{t + 1} - x_*\Vert^2
                & \quad 
                + \frac{1}{2}\left(
                    1 - (1 - (1 + \eta_{t + 1}\beta)^{-1})^2
                    \prod_{i = 1}^{t}
                        (1 + \eta_i \beta)^{-1}
                \right)\Vert x_0 - x_*\Vert^2. 
            $\phi$ is a $\eta_{t + 1}\beta$ strongly convex function, implying that 
                (1/2)\Vert x_{t} - x_*\Vert^2 \le (\eta_{t + 1}\beta)^{-1}(\phi(x_t) - \phi(x_*)). 
            Therefore at each iteration, the Lyapunov function would satisfy:

        \end{proof}
        \begin{remark}
            In both Guler\cite{guler_convergence_1991} and Rockafellar's \cite{rockafellar_monotone_1976} for PPM, they didn't mention the convergence of function value for strongly convex function explicitly. 
            The lack of interests could be due to the triviality of it, or, the assumption of strongly convex is too strong. 
            We do it here for completeness and pointing out the obvious.

@@

        \begin{example}[The fundamental proximal gradient lemma]
            The fundamental proximal gradient lemma was used heavily in the literature to derive convergence results in the convex case. 
            The "fundamental proximal gradient lemma" originates from Beck's writings \cite[theorem 10.16]{beck_first-order_nodate}. 
            We demonstrate in this example that it's a consequence of \hyperref[thm:ppm_breg_descent_ineq]{theorem \ref*{thm:ppm_breg_descent_ineq}}. 
            \par
            With $f = g + h$, $h$ convex, lsc, $g$ be $L$-Lipschitz smooth, then for all $y\in \RR^n$, $x\in \RR^n$, $y^+ := \hprox_{L^{-1}h}(y - L^{-1}\nabla (y))$ satisfies: 
            \begin{align*}
                f(x) - f(y^+) \ge \frac{L}{2}\Vert x - y^+\Vert^2 - \frac{L}{2}\Vert x - y\Vert^2 + D_g(x, y).
            \end{align*}
            A similar analysis as \hyperref[thm:lower_approx_ppm_convergence]{ theorem \ref*{thm:lower_approx_ppm_convergence}} with \hyperref[thm:ppm_breg_descent_ineq]{theorem \ref*{thm:ppm_breg_descent_ineq}} obtains the same inquality. 
            With $\phi(y) = \eta(h(y) + g(x) + \langle \nabla g(x), y - x\rangle)$ as an lower bounding function of $f$. 
            Let $x^+ = \hprox_\phi(x)$, then for all $u$:
            \begin{align*}
                & \phi(u) + \frac{1}{2}\Vert u - x\Vert^2 - \phi(x^+) - \frac{1}{2}\Vert x^+ - x\Vert^2 
                \ge \frac{1}{2}\Vert x^+ - u\Vert^2, 
                \\
                \implies &
                \eta\underbrace{(h(u) + g(x) + \langle \nabla g(x), u - x\rangle + \frac{L}{2}\Vert u - x\Vert^2)}_{\ge \phi(u)} 
                - \eta \underbrace{f(x^+)}_{\le \phi(x^+)} - \frac{1}{2}\Vert x - x^+\Vert^2 
                \\
                &\quad  
                + \left(
                    \frac{1}{2} - \frac{\eta L}{2}
                \right)\Vert u - x\Vert^2 
                \ge 
                \frac{1}{2}\Vert x^+ - u\Vert^2 
                \\
                \iff & 
                f(u) + g(x) - g(u) + 
                \langle \nabla g(x), u -x\rangle 
                - f(x^+) + \frac{L}{2}\Vert u - x\Vert^2 - \frac{1}{2\eta}\Vert x - x^+\Vert^2
                \\
                & \quad 
                + 
                \left(
                    \frac{1}{2\eta} - \frac{L}{2}
                \right)\Vert u - x\Vert^2 
                \ge 
                \frac{1}{2\eta}\Vert x^+ - u\Vert^2 
                \\
                \iff 
                & f(u) - f(x^+) - D_g(u, x)
                + \frac{1}{2\eta} \Vert u - x\Vert^2 - \frac{1}{2\eta}\Vert x^+ - x\Vert^2
                \ge 
                \frac{1}{2\eta} \Vert x^+ - u\Vert^2. 
            \end{align*}
            Removing the negative term $-1/2\eta \Vert x - x^+\Vert^2$ makes LHS larger, establishing the fundamental proximal gradient lemma. 
        \end{example}

@@

    \subsection{Varieties of Nesterov accelerated gradient}
        In this section, we list different varieties of the Nesterov accelerated method. 
        We present these varieties generically because these algorithms' forms are of interest.  
        \begin{definition}[Accelearted Gradient Generic Original Form]\label{def:agg_original}
            Let $f$ be a $L$ Lipschitz smooth and $\mu\ge 0$ strongly convex function. 
            Choose $x_0$, $\gamma_0 > 0$, set $v_0 = x_0$, for iteration $k\ge 0$, it
            \begin{enumerate}
                \item[1.] computes $\alpha_k \in (0, 1)$ by solving $L\alpha_k^2 = (1 - \alpha_k)\gamma_k + \alpha_k \mu$; 
                \item[2.] sets $\gamma_{k + 1} = (1 - \alpha_k)\gamma_k + \alpha_k \mu$;
                \item[3.] chooses $y_k = (\gamma_k + \alpha_k \mu)(\alpha_k \gamma_k v_k + \gamma_{k + 1}x_k)$. Compute $f(y_k)$ and $\nabla f(y_k)$; 
                \item[4.] finds $x_{k + 1}$ such that $f(x_{k + 1}) \le f(y_k) - (2L)^{-1} \Vert \nabla f(y_k)\Vert^2$; 
                \item[5.] sets $v_{k + 1} = \gamma_{k+1}^{-1}((1 - \alpha_k)\gamma_kv_k + \alpha_k \mu y_k - \alpha_k \nabla f(y_k))$. 
            \end{enumerate}
        \end{definition}
        \begin{remark}
            This is in Nesterov's book \cite[(2.2.7)]{nesterov_lectures_2018}. 
            It is the most generic algorithm in his book about accelerated gradient method. 
            The genericity of the algorithm is provided by item 4., which is the a special case of the smooth descent lemma. 
        \end{remark}

        \begin{definition}[Accelerated Gradient Generic Triangular Form]\label{def:agg_tri}

@@

        \begin{definition}[Accelerated Gradient Generic PPM Form]\label{agg_ppm}
